---
title: "Import Fric et al. results from supplemental tables for comparisons"
author: "Elise Larsen & Vaughn Shirey"
date: "Updated 12/7/2020; initiated 12/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```
<h2>*</h2>
  <p>Here we import Fric et al. results from supplemental tables.<br>
  We start with ~latitude and ~latitude|altitude+year model results </p>
```{r import latitude results from file}

##Needed packages:
library(readxl)
library(tidyverse)

##Import Fric results: 

#Check order of phenometric results in tables
#Single regression
fric.metrics1<-names(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~latitude", range="A3:Y3"))
fric.metrics1[grep("e",fric.metrics1)]
#Corrected regression (Regression of residuals)
fric.metrics2<-names(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~latitude|altitude+year", range="A2:Y2"))
fric.metrics2[grep("e",fric.metrics2)]

# Table is 1 species column and 3 sets of results columns (peak, onset, termination). There are 8 results parameters reported per model. 
n.metrics<-8
fric.results.lat<-na.omit(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~latitude", range="A4:Y113"))
#nrow(fric.results.lat)
fric.results.corr<-read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~latitude|altitude+year", range="A3:Y108")

fric.results.lat$model<-"lat"
fric.results.corr$model<-"corr"
fric.results<-rbind(fric.results.lat, fric.results.corr)

field.names<-c(names(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~latitude", range="B4:I4")))
field.names<-c("species",paste("peak",field.names,sep="."),paste("onset",field.names,sep="."),paste("term",field.names,sep="."),"model")

names(fric.results)<-field.names

```

<p> Add ~year and ~year|latitude responses for supplemental table 2</p>
 
```{r import year results}

#Check order of phenometric results in tables
#Single regression
yr.metrics1<-names(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~year", range="A2:Y2"))
yr.metrics1[grep("e",yr.metrics1)]
#Corrected regression (Regression of residuals)
yr.metrics2<-names(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~year|latitude", range="A2:Y2"))
yr.metrics2[grep("e",yr.metrics2)]

# Table is 1 species column and 3 sets of results columns (peak, onset, termination). There are 8 results parameters reported per model. 
n.metrics<-8
fric.yr<-na.omit(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~year", range="A3:Y108"))
#nrow(fric.results.lat)
fric.yr.lat<-read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~year|latitude", range="A3:Y108")

fric.yr$model<-"year"
fric.yr.lat$model<-"yearcorr"

field.names<-c(names(read_excel("fric_supplements/ele13419-sup-0003-tables2.xlsx", sheet="~year", range="B3:I3")))
field.names<-c("species",paste("peak",field.names,sep="."),paste("onset",field.names,sep="."),paste("term",field.names,sep="."),"model")

names(fric.yr.lat)<-names(fric.yr)<-field.names

#Combine all results into one table
fric.results<-rbind(fric.results,fric.yr,fric.yr.lat)
rm(fric.results.lat, fric.results.corr, fric.yr, fric.yr.lat)
rm(field.names, fric.metrics1, fric.metrics2, n.metrics, yr.metrics1, yr.metrics2)
#Keep Fields we want for comparison
fric.results<-fric.results %>%
  select(species,peak.ResDFMean,onset.ResDFMean,onset.p_mean,onset.coef,onset.response,term.p_mean,term.ResDFMean,term.coef,term.response, model)

```


```{r standardize names}
#standardize names
fric.results.species<-strsplit(fric.results$species," ")
result.names<-NULL
for(i in 1:length(fric.results.species)) {
  result.names<-c(result.names,paste(fric.results.species[[i]][1],fric.results.species[[i]][2],sep=" "))
}
fric.results$name<-result.names
#verify that names match
table(order(fric.results$name)-order(fric.results$species))
#The sort order is the same. Looks good.
```
<h2>*</h2>
<br> Because Fric et al. results are not tagged by region, we add a field for region.
We match results to region based on relative sample sizes.

```{r identify results by region}
#Get species-region & sample sizes from fric data file
#load occurrence data (see data curation files for how this file is created)

#loads fricdata 
load('data/occurrences_FricAnalysis.RData')
#summarize species-region-samplesize
datasummary<-fricdata %>%  group_by(name,region) %>% tally()

fric.results$region<-datasummary$region[match(fric.results$name,datasummary$name)]

#But now we need to correct the ones that are in both regions

#ID Species names that occur in both regions
fric.dups<-fric.results %>% group_by(name) %>% tally() %>% filter(n==8)

#Get data records with duplicates
dup.sp1<-datasummary[duplicated(datasummary$name) | duplicated(datasummary$name, fromLast=T),]

#For each species, check which region has more occurrence data
biggersample<-NULL
for(sp in sort(unique(dup.sp1$name))) {
  biggersample<-c(biggersample,ifelse(dup.sp1$n[dup.sp1$name==sp & dup.sp1$region=="Europe"]-dup.sp1$n[dup.sp1$name==sp & dup.sp1$region=="N. America"]>0,"Europe","N. America"))
}
table(biggersample)

#All Europe sample sizes are bigger so we can use that to ID regions in the Fric results.
for(sp in sort(unique(dup.sp1$name))) {
  sizes<-fric.results$peak.ResDFMean[which(fric.results$name==sp)]
  fric.results$region[fric.results$name==sp & fric.results$peak.ResDFMean==max(sizes)]<-"Europe"
  fric.results$region[fric.results$name==sp & fric.results$peak.ResDFMean==min(sizes)]<-"N. America"
}

fric.results$set<-paste(fric.results$name,fric.results$region,sep="-")

summary(fric.results)
print("Onset response: single regression")
table(filter(fric.results,model=="lat")$onset.response)
print("Termination response: single regression")
table(filter(fric.results,model=="lat")$term.response)
print("Onset response: residual regression")
table(filter(fric.results,model=="corr")$onset.response)
print("Termination response: residual regression")
table(filter(fric.results,model=="corr")$term.response)

#Save to data file
save(fric.results,file="data/fric_results.RData")
```
<br>The # responses in each group are correct.

<br>End File.