---
title: "Fric et al. Data formatting"
author: "Vaughn Shirey & Elise Larsen"
date: "Current version 3-Dec-2020; initiated Feb-2020"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#start fresh
rm(list=ls())
```
<h2>*</h2>

```{r load packages, message=F}

# load libraries
library(tidyverse)
library(ggplot2)
library(readxl)
library(lubridate)

```
<h2>*</h2>
<h3> Data Import and Formatting</h3>
data.csv file was downloaded from <a href="https://doi.org/10.6084/m9.figshare.9946934">https://doi.org/10.6084/m9.figshare.9946934</a>
(https://figshare.com/articles/Phenology_responses_of_temperate_butterflies_-_Supplementary_data/9946934)

<p>This cvs file contains the occurrence data used in Fric et al. (2020), which they downloaded from gbif. The file includes separate data tables for each dataset, which have been concatenated into one file. These data tables have the same fields but are not formatted as a single data table; individual datasets were all written into one data file, including headers and row indices in each dataset. This first set of code reformats the data & writes formatted data files. </p>


```{r format raw occurrence data table}
all.data <- readLines("fric_supplements/data.csv")

#identify header rows
all.header.rows<-grep("decimalLongitude", all.data)

#check headers for consistency
uniqueheaders<-unique(all.data[all.header.rows])

# 2 versions! 
#In the data curation file, we write the data for each header to a separate text file. See that file for the detailed code.
#read back in the formatted data
data1<-read_csv("data/fric_data_header_1.txt")
data2<-read_csv("data/fric_data_header_2.txt")
paste( nrow(data1), "records in format 1;", nrow(data2), "records in format 2")

alldata<-bind_rows(data1,data2)
rm(data1,data2)

##Fric et al includes different species names in results tables than found in data table. In the data curation folder, we match the data names to the results names and create the name_changes.csv file. Here we change names to match results tables:
names1<-read_csv("data/name_changes.csv")
# this file can now be used for correcting names in the main file

for(namei in 1:nrow(names1)) {
  alldata$name[alldata$name==names1$data.name[namei]]<-names1$result.name[namei]
}

rm(names1)

##Fric et al identifies datasets by region (N. America, Europe), but the data file does not include this information. We label data by region using longitude:
## visualize data density by longitude
#hist(alldata$decimalLongitude, main="Data density by Longitude") 
#We label everything East of -40 as Europe, the rest as N. America 
alldata<-alldata %>% 
  mutate(region=ifelse(decimalLongitude>=(-40),"Europe","N. America"))

#Fric et al removed all 1st of month observations and removed one species due to late season nests
fricdata<-filter(alldata, day!=1, name!="Euphydryas aurinia")

summary(fricdata)
#Save formatted and filtered occurrence data used by Fric et al. 
save(alldata,file="data/occurrences.RData")
save(fricdata,file="data/occurrences_FricAnalysis.RData")
```
<br>End of File.